---
title: "Assignment 6"
author: "Jakob Grøhn Damgaard"
date: "10/26/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(pacman)
p_load(tidyverse, gapminder, ggplot2, base, dplyr, randomcoloR)
```

# Question 1
**Define a defensive function that calculates the Gross Domestic Product of a nation from the data available in the gapminder dataset. Using that function, calculate the GDP of Denmark in the following years: 1967, 1977, 1987, 1997, 2007, and 2017.** 
\
```{r Question 1}
calcGDP <- function(data, year, country) { # input variables to the function is 'data', 'year' and 'country'
    data <- data[data$year %in% year, ] # subsetting data for specified year
    data <- data[data$country %in% country,] # subsetting data for specified country
  gdp <- data$pop * data$gdpPercap # calculating gdp 
  new <- cbind(data, gdp=gdp) # adding gdp as a column to the data frame
  return(new)
}
```
\
Running function on Denmark
```{r}
years <- unique(gapminder$year)[grep("7$", unique(gapminder$year),)] # defining a list of all years ending in '7'
for (year in years){ # looping through all the years in the list 'years' and running the function for each year
  row = calcGDP(gapminder, year, "Denmark")
  print(row) # printing output
}
```

# Question 2
**Write a script that loops over each country in the gapminder dataset, tests whether the country starts with a ‘B’ , and print out whether the life expectancy is smaller than 50, between 50 and 70, or greater than 70.**
\
```{r Quesrion 2}
for (country in unique(gapminder$country)){ # loops through each unique country name
  if (substring(country, 1, 1) == "B"){ # if-statement recogninsing if first letter in the country is 'B'
    country_b_subset = gapminder[gapminder$country %in% country,] # create a subset data for the specific country starting with a 'B'
    for (year in unique(country_b_subset$year)){ # loop through each year in the dataset 
      year_subset = country_b_subset[country_b_subset$year %in% year,] # create a subset of the countries data containing only data from the specific year
      if (year_subset$lifeExp < 50){ # checking life expectancy and printing a statement
        cat("In", year, "the country", country, "had a life expectancy below 50. \n")
      }
      else if (year_subset$lifeExp > 50 && year_subset$lifeExp < 70){
        cat("In", year, "the country", country, "had a life expectancy between 50 and 70. \n")
      }
      else{
        cat("In", year, "the country", country, "had a life expectancy above 70. \n")
      }
    }
  }
}
```
\
# Optional

**Write a script that loops over each country in the gapminder dataset, tests whether the country starts with a ‘M’ and graphs life expectancy against time (using plot() function) as a line graph if the mean life expectancy is under 50 years.**

```{r}
for (country in unique(gapminder$country)){ # loops through each unique country name
  if (substring(country, 1, 1) == "M"){ # if-statement recogninsing if first letter in the country is 'M'
    country_m_subset = gapminder[gapminder$country %in% country,] # create a subset data for the specific country starting with a 'M'
    colour = randomColor(luminosity = c("dark")) # setting a radnom colour to discriminate country-plots
    plot <- ggplot(country_m_subset, aes(x=year, y=lifeExp)) + # plotting line plot
      geom_line(colour = colour, size = 1.1)+
      theme_bw()+
      labs(x = "Year", # defining all labels
      y = "Life expectancy",
      title = country)
    print(plot) # printing plot
  }
}
```

