---
title: "Assignment 4 - Starting with R"
author: "Jakob Grøhn Damgaard"
date: "10/5/2020"
output: 
    html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Use R to figure out how many elements in the vector below are greater than 2 
Initially, I define the vector 'rooms'. I then proceed to discard all NAs using the 'na.omit()' function. Finally, I find the amount of elements with a value above two using the 'length()' function on the following subset of 'rooms':  rooms[room>2].
```{r 1)}
#Initializing the vector
rooms <- c(1, 2, 1, 3, 1, NA, 3, 1, 3, 2, 1, NA, 1, 8, 3, 1, 4, NA, 1, 3, 1, 2, 1, 7, 1, NA)

#Filtering out all NAs 
rooms <- na.omit(rooms)

#Counting number of elements with a value above 2
length(rooms[rooms>2])
```

The number of elements with a value above 2 is 8. 

## What is the result of running median() function on the above 'rooms' vector?

Running the 'median()' function on the na-omitted 'rooms' yields a median of 1.5
```{r 2)}
#Taking the median of the values in the vector
median(rooms)
```

## Download and read SAFI_clean.csv

```{r 3)}
#Installing packages
pacman::p_load(tidyverse)

#Creating subfolder for data
dir.create("data")

#Downloading data
download.file("https://ndownloader.figshare.com/files/11492171", "data/SAFI_clean.csv", mode = "wb")

# Reading data
interviews <- read_csv("data/SAFI_clean.csv")
```
